Generalized linear density ratio model (GLDRM) with frequency weights

### 2021-11-21 nuweb/wgldrm.tex
Add multiple importFrom to pass checks
Convert setup part of getTheta to C++
Complete C++ re-implementation of getTheta

@@

\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb4}{, 4}\NWlink{nuweb6}{, 6}\NWlink{nuweb10}{, 10}.

@@

@@

\mbox{}\verb@    logit <- thetaControl$logit@\\
\mbox{}\verb@    eps <- thetaControl$eps@\\
\mbox{}\verb@    maxiter <- thetaControl$maxiter@\\
\mbox{}\verb@    maxhalf <- thetaControl$maxhalf@\\
\mbox{}\verb@    maxtheta <- thetaControl$maxtheta@\\
\mbox{}\verb@    logsumexp <- thetaControl$logsumexp@\\

@@

\mbox{}\verb@    ## Value does not change@\\
\mbox{}\verb@    gMu <- g(mu, m, M)@\\
\mbox{}\verb@    ## Initialize values@\\
\mbox{}\verb@    theta <- thetaStart  # initial values required@\\
\mbox{}\verb@    thetaOld <- bPrimeErrOld <- rep(NA, n)@\\
\mbox{}\verb@    conv <- rep(FALSE, n)@\\
\mbox{}\verb@    maxedOut <- rep(FALSE, n)@\\
\mbox{}\verb@    if (logsumexp) {@\\
\mbox{}\verb@        logf0 <- log(f0)@\\
\mbox{}\verb@        logfUnstd <- logf0 + tcrossprod(spt, theta)@\\
\mbox{}\verb@        logb <- apply(logfUnstd, 2, logSumExp)@\\
\mbox{}\verb@        fTilt <- exp(logfUnstd - rep(logb, each=sptN))@\\
\mbox{}\verb@        normfact <- colSums(fTilt)@\\
\mbox{}\verb@        normss <- which(normfact != 1)@\\
\mbox{}\verb@        fTilt[, normss] <- fTilt[, normss, drop=FALSE] / rep(normfact[normss], each=sptN)@\\
\mbox{}\verb@        fUnstd <- f0 * exp(tcrossprod(spt, theta))  # |spt| x n matrix of tilted f0 values@\\
\mbox{}\verb@        b <- colSums(fUnstd)@\\
\mbox{}\verb@        fTilt <- fUnstd / rep(b, each=sptN)  # normalized@\\
\mbox{}\verb@    bPrime <- colSums(spt*fTilt)  # mean as a function of theta@\\
\mbox{}\verb@    bPrime2 <- colSums(outer(spt, bPrime, "-")^2 * fTilt)  # variance as a function of theta@\\
\mbox{}\verb@    bPrimeErr <- bPrime - mu  # used to assess convergence@\\
\mbox{}\verb@    ## Update theta until convergence@\\
\mbox{}\verb@    conv <- (abs(bPrimeErr) < eps) |@\\
\mbox{}\verb@        (theta==maxtheta & bPrimeErr<0) |@\\
\mbox{}\verb@        (theta==-maxtheta & bPrimeErr>0)@\\
\mbox{}\verb@    s <- which(!conv)@\\
\mbox{}\verb@    while(length(s)>0 && iter<maxiter) {@\\
\mbox{}\verb@        bPrimeErrOld[s] <- bPrimeErr[s]  # used to assess convergence@\\
\mbox{}\verb@        ## 1) Update theta@\\
\mbox{}\verb@        thetaOld[s] <- theta[s]@\\
\mbox{}\verb@        if (logit) {@\\
\mbox{}\verb@            tPrimeS <- (M-m) / ((bPrime[s]-m) * (M-bPrime[s])) * bPrime2[s]@\\
\mbox{}\verb@                # t'(theta) from paper: temporary variable@\\
\mbox{}\verb@                # only needed for the subset of observations that have not converged@\\
\mbox{}\verb@            tPrimeS[is.na(tPrimeS) | tPrimeS==Inf] <- 0@\\
\mbox{}\verb@                # If bPrime is on the boundary, then bPrime2 should be zero.@\\
\mbox{}\verb@                # Exceptions are due to rounding error.@\\
\mbox{}\verb@            thetaS <- theta[s] - (g(bPrime[s], m, M) - gMu[s]) / tPrimeS@\\
\mbox{}\verb@            thetaS <- theta[s] - bPrimeErr[s] / bPrime2[s]@\\
\mbox{}\verb@        thetaS[thetaS > maxtheta] <- maxtheta@\\
\mbox{}\verb@        thetaS[thetaS < -maxtheta] <- -maxtheta@\\
\mbox{}\verb@        theta[s] <- thetaS@\\
\mbox{}\verb@        ## 2) Update fTilt, bPrime, and bPrime2 and take half steps if bPrimeErr not improved@\\
\mbox{}\verb@        ss <- s@\\
\mbox{}\verb@        while(length(ss)>0 && nhalf<maxhalf) {@\\
\mbox{}\verb@            ## 2a) Update fTilt, bPrime, and bPrime2@\\
\mbox{}\verb@            if (logsumexp) {@\\
\mbox{}\verb@                logfUnstd[, ss] <- logf0 + tcrossprod(spt, theta[ss])@\\
\mbox{}\verb@                logb[ss] <- apply(logfUnstd[, ss, drop=FALSE], 2, logSumExp)@\\
\mbox{}\verb@                fTilt[, ss] <- exp(logfUnstd[, ss, drop=FALSE] - rep(logb[ss], each=sptN))@\\
\mbox{}\verb@                normfact <- colSums(fTilt[, ss, drop=FALSE])@\\
\mbox{}\verb@                normss <- which(normfact != 1)@\\
\mbox{}\verb@                fTilt[, ss[normss]] <- fTilt[, ss[normss], drop=FALSE] / rep(normfact[normss], each=sptN)@\\
\mbox{}\verb@                fUnstd[, ss] <- f0*exp(tcrossprod(spt, theta[ss]))  # |spt| x n matrix of tilted f0 values@\\
\mbox{}\verb@                b[ss] <- colSums(fUnstd[, ss, drop=FALSE])@\\
\mbox{}\verb@                fTilt[, ss] <- fUnstd[, ss, drop=FALSE] / rep(b[ss], each=sptN)  # normalized@\\
\mbox{}\verb@            bPrime[ss] <- colSums(spt*fTilt[, ss, drop=FALSE])  # mean as a function of theta@\\
\mbox{}\verb@            bPrime2[ss] <- colSums(outer(spt, bPrime[ss], "-")^2 * fTilt[, ss, drop=FALSE])  # variance as a function of theta@\\
\mbox{}\verb@            bPrimeErr[ss] <- bPrime[ss] - mu[ss]  # used to assess convergence@\\
\mbox{}\verb@            ## 2b) Take half steps if necessary@\\
\mbox{}\verb@            ss <- ss[abs(bPrimeErr[ss]) > abs(bPrimeErrOld[ss])]@\\
\mbox{}\verb@            if (length(ss) > 0) nhalf <- nhalf + 1@\\
\mbox{}\verb@            theta[ss] <- (theta[ss] + thetaOld[ss]) / 2@\\
\mbox{}\verb@        ## If maximum half steps are exceeded, set theta to previous value@\\
\mbox{}\verb@        maxedOut[ss] <- TRUE@\\
\mbox{}\verb@        theta[ss] <- thetaOld[ss]@\\
\mbox{}\verb@        ## 3) Check convergence@\\
\mbox{}\verb@        conv[s] <- (abs(bPrimeErr[s]) < eps) |@\\
\mbox{}\verb@            (theta[s]==maxtheta & bPrimeErr[s]<0) |@\\
\mbox{}\verb@            (theta[s]==-maxtheta & bPrimeErr[s]>0)@\\
\mbox{}\verb@        s <- s[!conv[s] & !maxedOut[s]]@\\

@@

\mbox{}\verb@    fTilt.extracted <- fTilt[cbind(ySptIndex, seq_along(ySptIndex))]@\\

@@

\mbox{}\verb@    list(theta=theta, fTilt=fTilt, bPrime=bPrime, bPrime2=bPrime2,@\\
\mbox{}\verb@         llik=llik, conv=conv, iter=iter)@\\

@@

\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb4}{, 4}\NWlink{nuweb6}{, 6}\NWlink{nuweb10}{, 10}.

@@

\NWtarget{nuweb10}{} \verb@"../R/wgldrm.R"@\nobreak\ {\footnotesize {10}}$\equiv$

>>>>

\mbox{}\verb@#' @{\tt @}\verb@importFrom stats lm.fit lm.wfit weighted.mean@\\

@@

\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb4}{, 4}\NWlink{nuweb6}{, 6}\NWlink{nuweb?}{, ?}.

@@

\mbox{}\verb@## usethis namespace: start@\\
\mbox{}\verb@#' @{\tt @}\verb@useDynLib exchreg, .registration = TRUE@\\
\mbox{}\verb@#' @{\tt @}\verb@importFrom Rcpp sourceCpp@\\
\mbox{}\verb@## usethis namespace: end@\\

@@

@@

\mbox{}\verb@    res <- getThetaC(spt, f0, mu, thetaStart, thetaControl)@\\

@@

\mbox{}\verb@    fTilt.extracted <- res$fTilt[cbind(ySptIndex, seq_along(ySptIndex))]@\\

@@

\mbox{}\verb@    list(theta=c(res$theta), fTilt=res$fTilt, bPrime=c(res$bPrime), bPrime2=c(res$bPrime2),@\\
\mbox{}\verb@         llik=llik, conv=c(res$conv), iter=res$iter)@\\

@@

\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb4}{, 4}\NWlink{nuweb6}{, 6}\NWlink{nuweb?}{, ?}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
The computationally intensive part of \texttt{getTheta} is implemented in C++. For now, only the \texttt{logsumexp = FALSE} and \texttt{logit = FALSE} settings are implemented.
\NWtarget{nuweb10}{} \verb@"../src/getTheta.cpp"@\nobreak\ {\footnotesize {10}}$\equiv$
\mbox{}\verb@#include <RcppArmadillo.h>@\\
\mbox{}\verb@using namespace Rcpp;@\\
\mbox{}\verb@using namespace arma;@\\
\mbox{}\verb@// [[Rcpp::depends(RcppArmadillo)]]@\\
\mbox{}\verb@// Computes log(sum(exp(x))) with better precision@\\
\mbox{}\verb@double logSumExp(arma::vec x){@\\
\mbox{}\verb@  uword i = x.index_max();@\\
\mbox{}\verb@  double m = x(i);@\\
\mbox{}\verb@  x.shed_row(i);@\\
\mbox{}\verb@  double lse = log1p(sum(exp(x-m))) + m;@\\
\mbox{}\verb@  return lse;@\\
\mbox{}\verb@// g function (logit transformation from appendix)@\\
\mbox{}\verb@arma::vec g(arma::vec mu, double m, double M){@\\
\mbox{}\verb@  arma::vec res = log(mu-m) - log(M-mu);@\\
\mbox{}\verb@  return res;@\\
\mbox{}\verb@}  @\\
\mbox{}\verb@// [[Rcpp::export]]@\\
\mbox{}\verb@List getThetaC(arma::vec spt, arma::vec f0, arma::vec mu,  @\\
\mbox{}\verb@               arma::vec thetaStart, List thetaControl){@\\
\mbox{}\verb@ // bool logit  = thetaControl["logit"];@\\
\mbox{}\verb@  double eps = thetaControl["eps"];@\\
\mbox{}\verb@  int maxiter = thetaControl["maxiter"];@\\
\mbox{}\verb@  double maxhalf = thetaControl["maxhalf"];@\\
\mbox{}\verb@  double maxtheta = thetaControl["maxtheta"];@\\
\mbox{}\verb@ // bool logsumexp = thetaControl["logsumexp"];@\\
\mbox{}\verb@  int sptN = spt.size();@\\
\mbox{}\verb@//  double m = min(spt);@\\
\mbox{}\verb@//  double M = max(spt);@\\
\mbox{}\verb@  int n = mu.size();@\\
\mbox{}\verb@  //initialize values@\\
\mbox{}\verb@  arma::vec theta = thetaStart;@\\
\mbox{}\verb@  arma::vec thetaOld(n);@\\
\mbox{}\verb@  arma::vec bPrimeErrOld(n);@\\
\mbox{}\verb@  uvec conv(n, fill::zeros);@\\
\mbox{}\verb@  uvec maxedOut(n, fill::zeros);@\\
\mbox{}\verb@  arma::rowvec oo(sptN, fill::ones);@\\
\mbox{}\verb@  mat fUnstd = exp(spt * theta.t());@\\
\mbox{}\verb@  fUnstd.each_col() %= f0;  // |spt| x n matrix of tilted f0 values@\\
\mbox{}\verb@  rowvec b = oo * fUnstd;  //column sums;@\\
\mbox{}\verb@  mat fTilt = fUnstd.each_row() / b;  // normalized@\\
\mbox{}\verb@  colvec bPrime = fTilt.t() * spt;  // mean as a function of theta@\\
\mbox{}\verb@  colvec bPrime2(n); // variance as a function of theta@\\
\mbox{}\verb@  for (int j=0; j<n; j++){       //iterate over the columns of fTilt@\\
\mbox{}\verb@    bPrime2(j) = 0;@\\
\mbox{}\verb@    for (int i=0; i<sptN; i++){  //iterate over the rows of fTilt@\\
\mbox{}\verb@      bPrime2(j) += pow(spt(i) - bPrime(j), 2.0) * fTilt(i,j);@\\
\mbox{}\verb@  colvec bPrimeErr = bPrime - mu;  // used to assess convergence@\\
\mbox{}\verb@  conv = (abs(bPrimeErr) < eps) || (theta==maxtheta && bPrimeErr<0) ||@\\
\mbox{}\verb@      (theta==-maxtheta && bPrimeErr>0);@\\
\mbox{}\verb@  uvec s = find(conv == 0);@\\
\mbox{}\verb@  int iter = 0;@\\
\mbox{}\verb@ while(s.size() > 0 && iter < maxiter) {@\\
\mbox{}\verb@    iter++;@\\
\mbox{}\verb@    bPrimeErrOld(s) = bPrimeErr(s);  // used to assess convergence@\\
\mbox{}\verb@// 1) Update theta@\\
\mbox{}\verb@    thetaOld(s) = theta(s);@\\
\mbox{}\verb@    colvec thetaS = theta(s) - bPrimeErr(s) / bPrime2(s);@\\
\mbox{}\verb@    thetaS(find(thetaS > maxtheta)).fill(maxtheta);@\\
\mbox{}\verb@    thetaS(find(thetaS < -maxtheta)).fill(-maxtheta);@\\
\mbox{}\verb@    theta(s)= thetaS;@\\
\mbox{}\verb@// 2) Update fTilt, bPrime, and bPrime2 and take half steps if bPrimeErr not improved@\\
\mbox{}\verb@    uvec ss = s;@\\
\mbox{}\verb@    int nhalf = 0;@\\
\mbox{}\verb@    while(ss.size() > 0 && nhalf < maxhalf) {@\\
\mbox{}\verb@// 2a) Update fTilt, bPrime, and bPrime2@\\
\mbox{}\verb@      fUnstd.cols(ss) = exp(spt * theta(ss).t());@\\
\mbox{}\verb@      fUnstd.each_col(ss) %= f0;  // |spt| x n matrix of tilted f0 values@\\
\mbox{}\verb@      b = oo * fUnstd.cols(ss);  //column sums;@\\
\mbox{}\verb@      mat tmp = fUnstd.cols(ss);@\\
\mbox{}\verb@      fTilt.cols(ss) = tmp.each_row() / b;  // normalized@\\
\mbox{}\verb@      bPrime(ss) = fTilt.cols(ss).t() * spt;  // mean as a function of theta@\\
\mbox{}\verb@      for (uword j=0; j<ss.size(); j++){       //iterate over the columns of fTilt[,ss]@\\
\mbox{}\verb@        bPrime2(ss(j)) = 0;@\\
\mbox{}\verb@        for (int i=0; i<sptN; i++){  //iterate over the rows of fTilt[,ss]@\\
\mbox{}\verb@          bPrime2(ss(j)) += pow(spt(i) - bPrime(ss(j)), 2.0) * fTilt(i,ss(j));@\\
\mbox{}\verb@      bPrimeErr(ss) = bPrime(ss) - mu(ss); @\\
\mbox{}\verb@// 2b) Take half steps if necessary@\\
\mbox{}\verb@      ss = ss(find(abs(bPrimeErr(ss)) > abs(bPrimeErrOld(ss))));@\\
\mbox{}\verb@      if (ss.size() > 0){@\\
\mbox{}\verb@        nhalf++;@\\
\mbox{}\verb@      } @\\
\mbox{}\verb@      theta(ss) = (theta(ss) + thetaOld(ss)) / 2;@\\
\mbox{}\verb@// If maximum half steps are exceeded, set theta to previous value@\\
\mbox{}\verb@    maxedOut(ss).fill(1);@\\
\mbox{}\verb@    theta(ss) = thetaOld(ss);@\\
\mbox{}\verb@// 3) Check convergence@\\
\mbox{}\verb@    conv(s) = (abs(bPrimeErr(s)) < eps) ||  (theta(s)==maxtheta && bPrimeErr(s) < 0) ||@\\
\mbox{}\verb@      (theta(s)==-maxtheta && bPrimeErr(s) > 0);@\\
\mbox{}\verb@    s = s(find(conv(s) == 0 && maxedOut(s) == 0));@\\
\mbox{}\verb@  List res;@\\
\mbox{}\verb@  res["theta"] = theta;@\\
\mbox{}\verb@  res["fTilt"] = fTilt;@\\
\mbox{}\verb@  res["bPrime"] = bPrime;@\\
\mbox{}\verb@  res["bPrime2"] = bPrime2;@\\
\mbox{}\verb@  res["conv"] = conv;@\\
\mbox{}\verb@  res["iter"] = iter;@\\
\mbox{}\verb@  return res;@\\

@@

\begin{flushleft} \small\label{scrap5}\raggedright\small
\NWtarget{nuweb?}{} \verb@"../R/wgldrm.R"@\nobreak\ {\footnotesize {?}}$\equiv$
